---
title: "faur ch 1 exercises"
output: pdf_document
date: "2025-08-03"
---

**1.1**
```{r 1.1}
x <- c(1, 2, 3)
y <- c(6, 5, 4)
```

Predict what will happen when the pieces of code are run.

a. `x * 2`
Each element in x is doubled
`2 4 6`
```{r 1.1.a}
x * 2
```

b. `x * y`
Each element in x is multiplied by the corresponding element in y
`6 10 12`
```{r 1.1.b}
x * y
```

c. `x[1] * y[2]`
First element of x is multiplied by second element of y
`5`
```{r 1.1.c}
x[1] * y[2]
```

**1.2** 
What is the value of `x` after each of the following commands?
a. `x + x`
`x` is still `1 2 3` because the new value of `x + x` is not reassigned to `x`

b. `x <- x + x`
`x` becomes `2 4 6` as each element in x gets incremented by itself (effectively doubled)

c. `y <- x + x`
`x` is still `1 2 3` because `x + x` is assigned to `y` not `x`

d. `x <- x + 1`
`x` becomes `2 3 4` as each element gets incremented by 1

**1.3** 
Determine the values of the vector `vec` after each of the following commands is run.

a. `vec <- 1:10`
`vec` becomes `1 2 3 4 5 6 7 8 9 10`

b. `vec <- 1:10 * 2`
`vec` becomes `2 4 6 8 10 12 14 16 18 20`

c. `vec <- 1:10^2`
`vec` becomes `1 2 3 4 ... 99 100`

d. `vec <- 1:10 + 1`
`vec` becomes `2 3 4 5 6 7 8 9 10 11`

e. `vec <- 1:(10 * 2)`
`vec` becomes `1 2 3 4 ... 19 20`

f. `vec <- rep(c(1,1,2), times = 2)`
`vec` becomes `1 1 2 1 1 2`

g. `vec <- seq(from = 0, to = 10, length.out = 5)`
`vec` becomes `0.0 2.5 5.0 7.5 10.0`

**1.4**
Graph the function $f(p) = p(1 - p)$ for $p \in [0,1]$.

a. Use `seq` to create a vector `p` of numbers from 0 to 1 spaced by 0.2
```{r 1.4.a}
p <- seq(from = 0, to = 1, by = 0.2)
```

b. Use `plot` to plot `p` in the `x` coordinate and `p(1 - p)` in the `y` coordinate.
```{r 1.4.b}
plot(x = p, y = p * (1 - p), type = "l")
```

c. Repeat, but with a vector `p` of numbers from 0 to 1 spaced by 0.01
```{r 1.4.c}
p <- seq(0, 1, by = 0.01)
plot(x = p, y = p * (1 - p), type = "l")
```

**1.5**
Use R to calculate the sum of the squares of all numbers from 1 to 100: $1^2 + 2^2 + \cdots 99^2 + 100^2$.
```{r 1.5}
sum((1:100)^2)
```

**1.6**
```{r 1.6}
x <- seq(from = 10, to = 30, by = 2)
```

a. What is the length of `x` (number of elements)?
```{r 1.6.a}
str(x)
length(x)
```

b. What is `x[2]`?
12

c. What is `x[1:5]`?
10 12 14 16 18

d. What is `x[1:3*2]`?
Equivalent to x[c(2, 4, 6)]
12 16 20

e. What is `x[1:(3*2)]`?
Equivalent to x[1:6]
10 12 14 16 18 20

f. What is `x > 25`?
FALSE x8 TRUE x3
```{r 1.6.f}
x > 25
```

g. What is `x[x > 25]`?
26 28 30

h. What is `x[-1]`?
12 14 16 18 20 22 24 26 28 30

i. What is `x[-1:-3]`
Equivalent to x[c(-1, -2, -3)]
16 18 20 22 24 26 28 30

**1.7**
R has built in vector `rivers` which contains the length of major North American rivers.

a. Use `?rivers` to learn about the dataset.
```{r 1.7.a}
?rivers
```
Contains 141 observations

b. Find the mean and standard deviation of the rivers data.
```{r 1.7.b}
# is.na(rivers)
mean(rivers)
sd(rivers)
```

c. Make a histogram (`hist`) of the rivers data.
```{r 1.7.c}
hist(rivers)
```

d. Get the five number summary (`summary`) of the rivers data.
```{r 1.7.d}
summary(rivers)
```

e. Find the longest and shortest length of rivers in the set.
```{r 1.7.e}
max(rivers)
min(rivers)
```

f. Make a list of all (lengths of) rivers longer than 1000 miles.
```{r 1.7.f}
rivers[rivers > 1000]
```

**1.8**
Consider the built-in data frame `airquality`.

a. How many observations of how many variables are there?
```{r 1.8.a}
str(airquality)
```
153 observations of 6 variables

b. What are the names of the variables?
Ozone, Solar.R, Wind, Temp, Month, Day

c. What type of data is each variable?
All are integers except Wind which is numeric

d. Do you agree with the data type given to each variable? What would have been some alternative choices?
Besides Month and Day, the rest could be numeric to capture more precise information

**1.9**
There is a built-in data set `state`, which is really seven separate variables with names such as `state.name`, `state.region`, and `state.area`.

a. What are the possible regions a state can be in? How many states are in each region?
```{r 1.9.a}
str(state.region)
levels(state.region)
table(state.region)
```

b. Which states have area less than 10,000 square miles?
```{r 1.9.b}
state.name[state.area < 10000]
```

c. Which state's geographic center is furthest south? (use `which.min`)
```{r 1.9.c}
str(state.center)
state.name[which.min(state.center$y)]
```

**1.10**
Consider the `mtcars` data set.

a. Which cars have 4 forward gears?
```{r 1.10.a}
str(mtcars)
mtcars[mtcars$gear == 4,]
```

b. What subset of `mtcars` does `mtcars[mtcars$disp > 150 & mtcars$mpg > 20,]` describe?
Cars with over 150 cubic inches of displacement and greater than 20 miles per gallon

c. Which cars have 4 forward gears and manual transmission? Note: manual transition is 1 and automatic is 0.
```{r 1.10.c}
mtcars[mtcars$gear == 4 & mtcars$am == 1,]
```

d. Which cars have 4 forward gears or manual transmission?
```{r 1.10.d}
mtcars[mtcars$gear == 4 | mtcars$am == 1,]
```

e. Find the mean mpg of the cars with two carburetors.
```{r 1.10.e}
mean(mtcars[mtcars$carb == 2, 1])
```

**1.11**
Consider the `mtcars` data set.

a. Convert the `am` variable to a factor with two levels, `auto` and `manual`.
```{r 1.11.a}
mtcars$am <- factor(mtcars$am, levels = c(0, 1), labels = c("auto", "manual"))
str(mtcars$am)
```

b. How many cars of each type of transmission are there?
```{r 1.11.b}
table(mtcars$am)
```

c. How many cars of each type of transmission have gas mileage estimates greater than 25 mpg?
```{r 1.11.c}
table(mtcars[mtcars$mpg > 25, "am"])
```

**1.12**
This problem uses the data set `hot_dogs` from the package `fosdata`.

a. How many kinds of observations are there? What types are the variables?
```{r 1.12.a}
str(fosdata::hot_dogs)
```

b. What are the three kinds of hot dogs in this data set?
```{r 1.12.b}
levels(fosdata::hot_dogs$type)
```

c. What is the highest sodium content of any hot dog in this data set?
```{r 1.12.c}
max(fosdata::hot_dogs$sodium)
```

d. What is the mean calorie content for Beef hot dogs?
```{r 1.12.d}
mean(fosdata::hot_dogs[fosdata::hot_dogs$type == "Beef", "calories"])
```

**1.13**
This problem uses the data set `DrinksWages` from the package `HistData`.
```{r 1.13 setup}
library(HistData)
```

a. How many observations of how many variables are there? What types are the variables?
```{r 1.13.a}
str(DrinksWages)
```

b. The variable `wage` contains the average wage for each profession. Which profession has the lowest wage?
```{r 1.13.b}
DrinksWages[which.min(DrinksWages$wage), "trade"]
```

c. The variables `n` contains the number of workers surveyed for each profession. Sum this to find the total number of workers surveyed.
```{r 1.13.c}
sum(DrinksWages$n)
```

d. Compute the mean wage for all workers surveyed by multiplying wage * n for each profession, summing, and dividing by the total number of workers surveyed.
```{r 1.13.d}
sum(DrinksWages$wage * DrinksWages$n) / sum(DrinksWages$n)
```

**1.14**
This problem uses the data set `Batting` from the `Lahman` package which contains batting statistics of all MLB players since 1871, broken down by season.
```{r 1.14 setup}
# install.packages("Lahman")
library(Lahman)
```

a. How many observations of how many variables are there?
```{r 1.14.a}
dim(Batting)
str(Batting)
```

b. Use the command `head(Batting)` to get a look at the first six lines of data.
```{r 1.14.b}
head(Batting)
```

c. What is the most number of triples (X3B) that have been hit in a single season?
```{r 1.14.c}
max(Batting$X3B)
```

d. What is the playerID(s) of the person(s) who hit the most number of triples in a single season? In what year did it happen?
```{r 1.14.d}
Batting[which.max(Batting$X3B), c("playerID", "yearID")]
```

e. Which player hit the most number of triples in a single season since 1960?
```{r 1.14.e}
People[People$playerID == Batting[Batting$yearID > 1960,][which.max(Batting[Batting$yearID > 1960, "X3B"]), "playerID"], c("nameFirst", "nameLast")]
```

**1.15**
Consider the `bechdel` data set in the `fosdata` package.
```{r 1.15}
str(fosdata::bechdel)
```

a. How many movies in the data set pass the Bechdel test?
```{r 1.15.a}
# nrow(fosdata::bechdel[fosdata::bechdel$binary == "PASS",])
sum(fosdata::bechdel$binary == "PASS")
```

b. What percentage of movies in the data set pass the Bechdel test?
```{r 1.15.b}
mean(fosdata::bechdel$binary == "PASS")
```

c. Create a table of number of movies in the data set by year.
```{r 1.15.c}
table(fosdata::bechdel$year)
```

d. Which year has the most movies in the data set?
```{r 1.15.d}
which.max(table(fosdata::bechdel$year))
```

e. How many different values are there in the `clean_test` variable?
```{r 1.15.e}
length(unique(fosdata::bechdel$clean_test))
```

f. Create a data frame that contains only those observations that pass the Bechdel test.
```{r 1.15.f}
pass <- fosdata::bechdel[fosdata::bechdel$binary == "PASS",]
rownames(pass) <- 1:nrow(pass) # reset index
head(pass)
```

g. Create a data frame that contains all of the observations that do not have missing values in the `domgross` variable.
```{r 1.15.g}
head(fosdata::bechdel[!is.na(fosdata::bechdel$domgross),])
```



